### 2014-10-09 Studies/2011_San_Francisco/Report/San_Francisco_Fire.tex
San Francisco: More edits based on reviewer comments

% PDF metadata
\hypersetup{
  pdftitle={Simulation of a Fire in a Hillside Residential Structure - San Francisco, CA},
  pdfauthor={Overholt, Kristopher J., Weinschenk, Craig G., and Madrzykowski, Daniel},
  pdfsubject={The Fire Dynamics Simulator (FDS) fire model, which is developed and maintained by the National Institute of Standards and Technology (NIST), was used to provide insight into the dynamics of a fire that occurred on June 2, 2011, within a multi-level, single-family residential structure in San Francisco, CA, that resulted in the death of two firefighters. The inputs for the FDS simulations are documented in this report and are based on the fire scenario, including the building geometry, interior furnishings, and ventilation conditions. The fire started in the basement and resulted in ventilation limited (fuel rich) fire conditions in the basement area. The fire originated in the basement, and the interior stairwell acted as a chimney for hot gases in the basement to flow towards areas of lower pressure via the interior stairwell towards the front of the structure. After the rear basement windows began to fail, the combustion gases at elevated temperature and pressure in the basement flowed upwards towards lower pressure areas via the stairwell. The temperature of the gases in the interior stairwell was estimated to be in excess of 700~$^{\circ}$C (1300~$^{\circ}$F). Two firefighters were located in the flow path between the basement and the doors on the front side of the structure, were exposed to these elevated temperatures, and later died as a result of their injuries.},
  pdfkeywords={flow path; fire dynamics; LODD; fire modeling; FDS simulation}
}

@@

The Fire Dynamics Simulator (FDS) fire model, which is developed and maintained by the National Institute of Standards and Technology (NIST), was used to provide insight into the dynamics of a fire that occurred on June 2, 2011, within a multi-level, single-family residential structure in San Francisco, CA, that resulted in the death of two firefighters. The inputs for the FDS simulations are documented in this report and are based on the fire scenario, including the building geometry, interior furnishings, and ventilation conditions. The fire started in the basement and resulted in ventilation limited (fuel rich) fire conditions in the basement area.

@@

On June 2, 2011, a fire in a multi-level, wood-frame residential structure claimed the life of two firefighters of the San Francisco Fire Department. NIST examined the fire dynamics of this incident at the request of the National Institute for Occupational Safety and Health (NIOSH) and the San Francisco Fire Department. Computer simulations of the fire incident were conducted using the NIST Fire Dynamics Simulator~\cite{FDS_Users_Guide} and Smokeview~\cite{Smokeview_Users_Guide} software to provide insight into the fire development and thermal conditions that likely existed in the residence during the fire. The specific objectives of the simulations detailed in this report are: 
\begin{enumerate}
\item To examine the effect of fire-induced flow paths (including temperature, pressure, and fire conditions) in this multi-level residential structure using physics-based calculations.
\item To provide visualizations of the fire behavior that are representative of the conditions that members of the San Francisco Fire Department likely experienced during the course of their interior operations.
\end{enumerate}
This document describes the input and the results of the FDS (version 6.1.2) simulations. The simulations were developed using a combination of knowledge of the fire scenario and appropriate engineering approximations and assumptions. Analysis of the simulation results focuses on the hazardous conditions that developed following the development of a flow path inside of the structure. This document is organized as follows: Section~\ref{sec:incident_summary} provides a summary of the fire incident, Section~\ref{sec:model_description} describes the relevant model input parameters and assumptions that were used to develop the simulations, Section~\ref{sec:model_results} presents the simulation results, and Section~\ref{sec:discussion} discusses the simulation results as they relate to firefighter safety and effectiveness. Appendix~\ref{sec:drawings} contains dimensioned drawings of the basement, first, and second floors of the structure.

@@

The basement primarily contained large upholstered polyurethane furniture items, and the balcony was constructed of wood. The HRR of the initial couch fire was based on an experimentally measured HRR for a mattress fire~\cite{madrzykowski2009fire} to represent the fire growth rate of a similar household furniture item. The HRRs of the additional furniture items were estimated based on previous experimental work that characterized the HRR of similar upholstered furniture fuel items~\cite{madrzykowski2009fire,Madrzykowski:2,Babrauskas:1}. Other combustible items and furnishings such as tables, stools, bookshelves, and carpet were present in the basement living room. However, to simplify the analysis and specification of the design fire, only the large upholstered furniture fuel items and balcony were included in the simulation because they were considered to have contributed the largest amount of energy in this fire scenario.

@@

In reality, the quantities associated with model input parameters are not fixed values; rather, a model input parameter can be thought of as a point estimate from a distribution of possible input parameters with some associated amount of uncertainty. Any change in an input parameter (such as the HRR) for a given scenario would result in a change in the output quantity (such as the hot gas layer (HGL) temperature). For example, according to the McCaffrey, Quintiere, and Harkleroad~\cite{SFPE:Walton} empirical correlation, the HGL temperature in a well-ventilated compartment fire is proportional to the HRR raised to the two-thirds power. Following this relationship, a 7.5~\% increase in the HRR would result in a 5~\% increase in the HGL temperature~\cite{NUREG_1824_Sup_1}. More detailed discussion on the propagation of parameter uncertainty in fire models is available in a validation study that was sponsored by the U.S. Nuclear Regulatory Commission~\cite{NUREG_1824_Sup_1}. The following subsections describe the inputs that were used to develop the simulation, including any approximations or assumptions that were made.

@@

As discussed in Section~\ref{sec:geometry}, the structure had exterior dimensions of 8.5~m (28~ft) by 17.7~m (58~ft) and a height of 5.9~m (19~ft) above street level. To ensure adequately resolved fluid flow in and out of the structure, the computational domain was extended beyond the volume of the structure. The dimensions of the computational domain were 11.5~m (38~ft) by 22~m (72~ft) and a height of 13.6~m (45~ft). A grid resolution of 10~cm (3.9~in) was used, which resulted in a total of approximately 3.4 million computational cells. As a result, all of the input model geometry lengths, ventilation openings (holes, doors, and windows), and fire areas snapped to the nearest 10~cm. Figure~\ref{fig:SMV_exterior_grid} shows the front and rear sides of the structure rendered in Smokeview with the 10~cm mesh.

@@

The domain was divided into 16 meshes (each containing between 184,000 and 248,200 grid cells) that could be processed in parallel, which reduced the amount of required calculation time to approximately 2~days. Figure~\ref{fig:mult_mesh} shows the entire structure within the computational domain that has been divided into multiple meshes. Table~\ref{tab:model_parameters} shows a summary of the model input parameters for the simulations that were conducted as part of this study.

@@

@@

To examine the results of the simulations, it is important to link the timelines from the fire scene to the simulation times. Table~\ref{tab:combined_timeline} shows the fireground timeline~\cite{NIOSH:Bowyer2} along with the corresponding simulation times.

@@

The results of the simulations are discussed in the following sections. Section~\ref{sec:simulated_flow_path} addresses the results of the simulation as they relate to the flow path that was established in the interior stairwell. Section~\ref{sec:assessing_the_hazard} examines the model results as they relate to the hazardous conditions and exposure temperatures in the interior stairwell. Section~\ref{sec:tactical_condsiderations} discusses tactical considerations and outcomes of the fire incident as they relate to the fire dynamics and flow path conditions that have been observed in recent experimental research.

@@

NIST's Fire Dynamics Simulator was used to provide insight into the fire dynamics of a fire that occurred within a multi-level, single-family residential structure in San Francisco, CA, that resulted in the death of two firefighters. The fuel, fire size, and fire growth rate that were used in the FDS simulations were estimated by taking into account all of the available information including the NIOSH post-incident report~\cite{NIOSH:Bowyer2}, post-incident pictures, and relevant literature. This resulted in a maximum specified source fire of approximately 20~MW in the basement and rear balcony of the structure. Based on the limited ventilation conditions in the basement, the FDS simulation results indicated a HRR of approximately 2~MW as a result of the limited supply of oxygen. After the rear basement windows failed, the HRR increased to approximately 32~MW before it reached a steady-state value of approximately 20~MW.

>>>>

% PDF metadata
\hypersetup{
  pdftitle={Simulation of a Fire in a Hillside Residential Structure - San Francisco, CA},
  pdfauthor={Overholt, Kristopher J., Weinschenk, Craig G., and Madrzykowski, Daniel},
  pdfsubject={The Fire Dynamics Simulator (FDS) fire model, which is developed and maintained by the National Institute of Standards and Technology (NIST), was used to provide insight into the dynamics of a fire that occurred on June 2, 2011, within a multi-level, single-family residential structure in San Francisco, CA, that resulted in the death of two firefighters. The inputs for the FDS simulation are documented in this report and are based on the fire scenario, including the building geometry, interior furnishings, and ventilation conditions. The fire started in the basement and resulted in ventilation limited (fuel rich) fire conditions in the basement area. The fire originated in the basement, and the interior stairwell acted as a chimney for hot gases in the basement to flow towards areas of lower pressure via the interior stairwell towards the front of the structure. After the rear basement windows began to fail, the combustion gases at elevated temperature and pressure in the basement flowed upwards towards lower pressure areas via the stairwell. The temperature of the gases in the interior stairwell was estimated to be in excess of 700~$^{\circ}$C (1300~$^{\circ}$F). Two firefighters were located in the flow path between the basement and the doors on the front side of the structure, were exposed to these elevated temperatures, and later died as a result of their injuries.},
  pdfkeywords={flow path; fire dynamics; LODD; fire modeling; FDS simulation}
}

@@

The Fire Dynamics Simulator (FDS) fire model, which is developed and maintained by the National Institute of Standards and Technology (NIST), was used to provide insight into the dynamics of a fire that occurred on June 2, 2011, within a multi-level, single-family residential structure in San Francisco, CA, that resulted in the death of two firefighters. The inputs for the FDS simulation are documented in this report and are based on the fire scenario, including the building geometry, interior furnishings, and ventilation conditions. The fire started in the basement and resulted in ventilation limited (fuel rich) fire conditions in the basement area.

@@

On June 2, 2011, a fire in a multi-level, wood-frame residential structure claimed the life of two firefighters of the San Francisco Fire Department. NIST examined the fire dynamics of this incident at the request of the National Institute for Occupational Safety and Health (NIOSH) and the San Francisco Fire Department. A computer simulation of the fire incident was conducted using the NIST Fire Dynamics Simulator~\cite{FDS_Users_Guide} and Smokeview~\cite{Smokeview_Users_Guide} software to provide insight into the fire development and thermal conditions that likely existed in the residence during the fire. The specific objectives of the simulation detailed in this report are:
\begin{enumerate}
\item To examine the effect of fire-induced flow paths (including temperature, pressure, and fire conditions) in this multi-level residential structure using physics-based calculations.
\item To provide visualizations of the fire behavior that are representative of the conditions that members of the San Francisco Fire Department likely experienced during the course of their interior operations.
\end{enumerate}
This document describes the input and the results of the FDS (version 6.1.2) simulation. The simulation was developed using a combination of knowledge of the fire scenario and appropriate engineering approximations and assumptions. Analysis of the simulation results focuses on the hazardous conditions that developed following the development of a flow path inside of the structure. This document is organized as follows: Section~\ref{sec:incident_summary} provides a summary of the fire incident, Section~\ref{sec:model_description} describes the relevant model input parameters and assumptions that were used to develop the simulation, Section~\ref{sec:model_results} presents the simulation results, and Section~\ref{sec:discussion} discusses the simulation results as they relate to firefighter safety and effectiveness. Appendix~\ref{sec:drawings} contains dimensioned drawings of the basement, first, and second floors of the structure.

@@

The basement primarily contained upholstered polyurethane furniture, and the balcony was constructed of wood. The upholstered furniture items located in the middle of the basement living room were very large and flat, similar to the size and geometry of a mattress. Therefore, the HRR of the initial couch fire was based on an experimentally measured HRR for a mattress fire~\cite{madrzykowski2009fire} to represent the fire growth rate of a similar household furniture item. The HRRs of the additional furniture items were estimated based on previous experimental work that characterized the HRR of similar upholstered furniture fuel items~\cite{madrzykowski2009fire,Madrzykowski:2,Babrauskas:1}. Other combustible items and furnishings such as tables, stools, bookshelves, and carpet were present in the basement living room. However, to simplify the analysis and specification of the design fire, only the large upholstered furniture fuel items and balcony were included in the simulation because they were considered to have contributed the largest amount of energy in this fire scenario.

@@

@@

As discussed in Section~\ref{sec:geometry}, the structure had exterior dimensions of 8.5~m (28~ft) by 17.7~m (58~ft) and a height of 5.9~m (19~ft) above street level. To ensure adequately resolved fluid flow in and out of the structure, the computational domain was extended beyond the volume of the structure. The dimensions of the computational domain were 11.5~m (38~ft) by 22~m (72~ft) and a height of 13.6~m (45~ft). A grid resolution of 10~cm (3.9~in) was used, which resulted in a total of approximately 3.4 million computational cells. As a result, all of the obstructions, ventilation openings (holes, doors, and windows), and fire areas snapped to the nearest 10~cm. Figure~\ref{fig:SMV_exterior_grid} shows the front and rear sides of the structure rendered in Smokeview with the 10~cm grid cells.

@@

The domain was divided into 16 meshes (each containing between 184,000 and 248,200 grid cells) that could be processed in parallel, which reduced the amount of required calculation time to approximately 2~days. Figure~\ref{fig:mult_mesh} shows the entire structure within the computational domain that has been divided into multiple meshes.

@@

\section{Summary of Model Input Parameters}
\label{sec:summary_of_model_inputs}

Table~\ref{tab:model_parameters} shows a summary of the model input parameters for the simulation that was conducted as part of this study. In reality, the quantities associated with model input parameters are not fixed values; rather, a model input parameter can be thought of as a point estimate from a distribution of possible input parameters with some associated amount of uncertainty. Any change in an input parameter (such as the HRR) for a given scenario results in a change in the output quantity (such as the hot gas layer (HGL) temperature).

For example, according to the McCaffrey, Quintiere, and Harkleroad~\cite{SFPE:Walton} empirical correlation, the HGL temperature in a well-ventilated compartment fire is proportional to the HRR raised to the two-thirds power. Following this relationship, a 7.5~\% increase in the HRR would result in a 5~\% increase in the HGL temperature~\cite{NUREG_1824_Sup_1}. More detailed discussion on the propagation of parameter uncertainty in fire models is available in a validation study that was sponsored by the U.S. Nuclear Regulatory Commission~\cite{NUREG_1824_Sup_1}.

@@

To examine the results of the simulation, it is important to link the timelines from the fire scene to the simulation times. Table~\ref{tab:combined_timeline} shows the fireground timeline~\cite{NIOSH:Bowyer2} along with the corresponding simulation times.

@@

The results of the simulation are discussed in the following sections. Section~\ref{sec:simulated_flow_path} addresses the results of the simulation as they relate to the flow path that was established in the interior stairwell. Section~\ref{sec:assessing_the_hazard} examines the model results as they relate to the hazardous conditions and exposure temperatures in the interior stairwell. Section~\ref{sec:tactical_condsiderations} discusses tactical considerations and outcomes of the fire incident as they relate to the fire dynamics and flow path conditions that have been observed in recent experimental research.

@@

NIST's Fire Dynamics Simulator was used to provide insight into the fire dynamics of a fire that occurred within a multi-level, single-family residential structure in San Francisco, CA, that resulted in the death of two firefighters. The fuel, fire size, and fire growth rate that were used in the FDS simulation were estimated by taking into account all of the available information including the NIOSH post-incident report~\cite{NIOSH:Bowyer2}, post-incident pictures, and relevant literature. This resulted in a maximum specified source fire of approximately 20~MW in the basement and rear balcony of the structure. Based on the limited ventilation conditions in the basement, the FDS simulation results indicated a HRR of approximately 2~MW as a result of the limited supply of oxygen. After the rear basement windows failed, the HRR increased to approximately 32~MW before it reached a steady-state value of approximately 20~MW.
